---
title: "Analysis and modelling on the Capital Bikeshare data"
author: "[Anusha Umashankar](https://github.com/Anusha-raju)| [Aidan Carlisle](https://github.com/acarlisle8)| [Rachel Thomas](https://github.com/releered)| [Sayan Patra](https://github.com/Sayanpatraa) - Group 8 "
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<style>

h1 { color: #800000; }  

h2 { color: #000080; } 
h3 { color: #004d4d; } 

</style>

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
```

```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(geosphere)
library(tidygeocoder)
library(caret)
library(lattice)
library(lubridate)
library(rattle)
library(rpart)
library(rpart.plot)
library(car)
library(pROC)
library(rattle)
library(caret) 
```


```{r}
file_path <- file.path(getwd(), "202410-capitalbikeshare-tripdata.csv")
capital_bikeshare <- read.csv(file_path, header = TRUE)
```

The dataset contains the following attributes.
```{r}
colnames(capital_bikeshare)
```

# Preprocessing:

1. There are data points where information like start_station_name & id , end_station_name & id , and end_lat and end_lng are missing. 
`r colSums(is.na(capital_bikeshare))`.

Dropping the `r colSums(is.na(capital_bikeshare))["end_lat"]` rows as they consistute only `r colSums(is.na(capital_bikeshare))["end_lat"]/nrow(capital_bikeshare)*100` percent of the dataset. 

```{r}
capital_bikeshare <- capital_bikeshare %>%
  filter(!is.na(end_lat))
```
2. `Distance (in miles)` is derived using the latitude and longitude coordinates of the start and end points. The `distVincentySphere()` function from the `geosphere` package in R is used to compute the geodesic distance, factoring in the Earth's curvature.

```{r}
# Calculate the distance for each row and add it as a new column (in miles)
capital_bikeshare$distance_miles <- mapply(function(lat1, lon1, lat2, lon2) {
  distVincentySphere(c(lon1, lat1), c(lon2, lat2)) / 1000 * 0.621371  
}, capital_bikeshare$start_lat, capital_bikeshare$start_lng, capital_bikeshare$end_lat, capital_bikeshare$end_lng)
```
3. Calculated the total ride time by subtracting `started_at` from `ended_at.`
```{r}
capital_bikeshare$started_at <- as.POSIXct(capital_bikeshare$started_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ended_at <- as.POSIXct(capital_bikeshare$ended_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ride_duration_mins <- as.numeric(difftime(capital_bikeshare$ended_at, capital_bikeshare$started_at, units = "mins"))
```
4. Categorizing Rides by Time Period.
```{r}
get_time_period <- function(time) {
  hour <- as.numeric(format(time, "%H"))
  
  if (hour >= 4 & hour < 6) {
    return("Early Morning")
  } else if (hour >= 6 & hour < 9) {
    return("Morning")
  } else if (hour >= 9 & hour < 12) {
    return("Mid-Morning")
  } else if (hour >= 12 & hour < 13) {
    return("Midday")
  } else if (hour >= 13 & hour < 17) {
    return("Afternoon")
  } else if (hour >= 17 & hour < 19) {
    return("Evening")
  } else if (hour >= 19 & hour < 21) {
    return("Late Evening")
  } else {
    return("Night")
  }
}


capital_bikeshare <- capital_bikeshare %>%
  mutate(
    start_time_period = sapply(started_at, get_time_period),
  )
```
5. Categorizing rides by weekday vs weekend
```{r}
capital_bikeshare$day_type <- ifelse(wday(capital_bikeshare$started_at) %in% c(1, 7), "weekend", "weekday")

```


**The following attributes are considered for further Exploratory Data Analysis (EDA):**

```{r}
filtered_capital_bikeshare <- capital_bikeshare %>% select(rideable_type,member_casual,distance_miles,ride_duration_mins, start_time_period,day_type)
```

*Factorizing 'rideable_type', 'member_casual', 'start_time_period', and 'day_type' variables:*
```{r}
filtered_capital_bikeshare$rideable_type <- as.factor(filtered_capital_bikeshare$rideable_type) 
filtered_capital_bikeshare$member_casual <- as.factor(filtered_capital_bikeshare$member_casual)
filtered_capital_bikeshare$start_time_period <- as.factor(filtered_capital_bikeshare$start_time_period)
filtered_capital_bikeshare$day_type <- as.factor(filtered_capital_bikeshare$day_type)
```

*Removing outliers*
```{r}
#for distance_miles
z_scores_DM <- scale(filtered_capital_bikeshare$distance_miles)
outliers <- which(abs(z_scores_DM) > 3)
capital_bikeshare_clean <- filtered_capital_bikeshare[-outliers, ]

# for ride_duration_mins
z_scores_RDM <- scale(capital_bikeshare_clean$ride_duration_mins)
outliers <- which(abs(z_scores_RDM) > 2)
capital_bikeshare_clean <- capital_bikeshare_clean[-outliers, ]

```

After cleaning the dataset the number of rows are `r nrow(capital_bikeshare_clean)`.

# Visualizations:
## 1. Frequency of electric bike versus classic bike rentals
```{r}

bike_type_counts <- table(capital_bikeshare_clean$rideable_type)
bike_type_df <- as.data.frame(bike_type_counts)

ggplot(bike_type_df, aes(x = Var1, y = Freq, fill = Var1)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Frequency of Electric Bike vs Classic Bike Rentals",
       x = "Bike Type",
       y = "Frequency of Rentals") + 
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "green")) +  
  scale_y_continuous(labels = scales::comma)+
  theme(legend.position = "none")
```

- The higher frequency of electric bike rentals compared to classic bikes in the dataset indicates a preference for electric bikes. 

## 2. Distribution of duration rides on electric bike versus classic bike

*Histogram*
```{r}
ggplot(capital_bikeshare_clean, aes(x = ride_duration_mins, fill = rideable_type)) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.7, position = "dodge") +
  labs(
    title = "Distribution of Ride Duration by Bike Type",
    x = "Ride Duration (Minutes)",
    y = "Frequency"
  ) +
  scale_fill_manual(values = c("electric_bike" = "green", "classic_bike" = "blue")) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```

*Box plot*
```{r}
ggplot(capital_bikeshare_clean, aes(x = rideable_type, y = ride_duration_mins, fill = rideable_type)) +
  geom_boxplot() +
  labs(title = "Boxplot of Ride Durations for Electric vs Classic Bikes",
       x = "Bike Type",
       y = "Ride Duration (minutes)") +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "green")) +
  theme_minimal()
```

- The distribution of ride duration by bike type is right-skewed, and the box plot highlights a significant number of extreme outliers.


## Distribution of distances ride by bike type. 

*Histogram*
```{r}
ggplot(capital_bikeshare_clean, aes(x = distance_miles, fill = rideable_type)) +
  geom_histogram(binwidth = 0.3, color = "black", alpha = 0.7, position = "dodge") +
  labs(
    title = "Distribution of Ride Distance by Bike Type",
    x = "Ride Distance (miles)",
    y = "Frequency"
  ) +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "green")) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```
*Box-plot*
```{r}
ggplot(capital_bikeshare_clean, aes(x = rideable_type, y = distance_miles, fill = rideable_type)) +
  geom_boxplot() +
  labs(
    title = "Box Plot of Ride Distance by Bike Type",
    x = "Rideable Type",
    y = "Distance (miles)"
  ) +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "green")) + 
  theme_minimal()
```


- The distribution of ride distance by bike type is slightly more right-skewed for classic bikes compared to electric bikes, with electric bikes exhibiting a wider distribution than classic bikes.


## Frequency of user types
```{r}
user_type_counts <- table(capital_bikeshare_clean$member_casual)
user_type_df <- as.data.frame(user_type_counts)
ggplot(user_type_df, aes(x = "", y = Freq, fill = User_Type)) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar(theta = "y") + 
  labs(title = "Proportion of User Types (Member vs Casual)") +
  scale_fill_manual(values = c("member" = "blue", "casual" = "green")) +
  theme_void() +
  geom_text(aes(label = paste0(round(Freq/sum(Freq)*100, 1), "%")), 
            position = position_stack(vjust = 0.5), color = "black") 
```

- There is a significantly higher number of member-type riders in October.



## User Types by Rideable Type

```{r}
user_bike_counts <- capital_bikeshare_clean %>%
  group_by(member_casual, rideable_type) %>%
  summarise(Freq = n())  

ggplot(user_bike_counts, aes(x = rideable_type, y = Freq, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "User Type Distribution by Rideable Type",
       x = "Rideable Type (Bike Type)",
       y = "Frequency of Rides") +
  scale_fill_manual(values = c("member" = "blue", "casual" = "green")) +  
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```


- Electric bikes have a higher proportion of member-type riders compared to casual-type riders.



