---
title: "Analysis and modelling on the Capital Bikeshare data"
author: "[Anusha Umashankar](https://github.com/Anusha-raju)| [Aidan Carlisle](https://github.com/acarlisle8)| [Rachel Thomas](https://github.com/releered)| [Sayan Patra](https://github.com/Sayanpatraa) - Group 8 "
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<style>

h1 { color: #800000; }  

h2 { color: #000080; } 
h3 { color: #004d4d; } 

</style>

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(geosphere)
library(tidygeocoder)
library(caret)
library(lattice)
library(lubridate)
library(rattle)
library(rpart)
library(rpart.plot)
library(car)
library(pROC)
```



capital_bikeshare <- X202410_capitalbikeshare_tripdata


```{r}
file_path <- file.path(getwd(), "X202410_capitalbikeshare_tripdata.csv")
capital_bikeshare <- read.csv(file_path, header = TRUE)
```

The dataset contains the following attributes.
```{r}
colnames(capital_bikeshare)
```

# Preprocessing:

1. 
There are data points where information like start_station_name & id , end_station_name & id , and end_lat and end_lng are missing. 
`r colSums(is.na(capital_bikeshare))`.

Dropping the `r colSums(is.na(capital_bikeshare))["end_lat"]` rows as they consitute only `r colSums(is.na(capital_bikeshare))["end_lat"]/nrow(capital_bikeshare)*100` percent of the dataset. 

```{r}
capital_bikeshare <- capital_bikeshare %>%
  filter(!is.na(end_lat))
```


2. Reverse Geocoding: Converting Latitude and Longitude Coordinates to Addresses
```{r}
#capital_bikeshare$start_addy<-reverse_geocode(capital_bikeshare, lat = "start_lat", long = "start_lng")
#capital_bikeshare$end_addy<-reverse_geocode(capital_bikeshare, lat = "end_lat", long = "end_lng")
```

3. `Distance (in miles)` is derived using the latitude and longitude coordinates of the start and end points. The `distVincentySphere()` function from the `geosphere` package in R is used to compute the geodesic distance, factoring in the Earth's curvature.

```{r}
# Calculate the distance for each row and add it as a new column (in miles)
capital_bikeshare$distance_miles <- mapply(function(lat1, lon1, lat2, lon2) {
  distVincentySphere(c(lon1, lat1), c(lon2, lat2)) / 1000 * 0.621371  
}, capital_bikeshare$start_lat, capital_bikeshare$start_lng, capital_bikeshare$end_lat, capital_bikeshare$end_lng)
```


4. Calculated the total ride time by subtracting `started_at` from `ended_at.`
```{r}
capital_bikeshare$started_at <- as.POSIXct(capital_bikeshare$started_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ended_at <- as.POSIXct(capital_bikeshare$ended_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ride_duration_mins <- as.numeric(difftime(capital_bikeshare$ended_at, capital_bikeshare$started_at, units = "mins"))
```


5. Categorizing Rides by Time Period.
```{r}
get_time_period <- function(time) {
  hour <- as.numeric(format(time, "%H"))
  
  if (hour >= 4 & hour < 6) {
    return("Early Morning")
  } else if (hour >= 6 & hour < 9) {
    return("Morning")
  } else if (hour >= 9 & hour < 12) {
    return("Mid-Morning")
  } else if (hour >= 12 & hour < 13) {
    return("Midday")
  } else if (hour >= 13 & hour < 17) {
    return("Afternoon")
  } else if (hour >= 17 & hour < 19) {
    return("Evening")
  } else if (hour >= 19 & hour < 21) {
    return("Late Evening")
  } else {
    return("Night")
  }
}


capital_bikeshare <- capital_bikeshare %>%
  mutate(
    start_time_period = sapply(started_at, get_time_period),
  )
```

6. Categorizing rides by weekday vs weekend
```{r}
capital_bikeshare$day_type <- ifelse(wday(capital_bikeshare$started_at) %in% c(1, 7), "weekend", "weekday")

```



**The following attributes are considered for further Exploratory Data Analysis (EDA):**

```{r}
#filtered_capital_bikeshare <- capital_bikeshare %>% select(rideable_type,member_casual,distance_miles,ride_duration_mins, start_time_period, start_addy, end_addy)
filtered_capital_bikeshare <- capital_bikeshare %>% select(rideable_type,member_casual,distance_miles,ride_duration_mins, start_time_period,day_type)
```

*Factorizing 'rideable_type', 'member_casual', 'start_time_period', and 'day_type' variables:*
```{r}
filtered_capital_bikeshare$rideable_type <- as.factor(filtered_capital_bikeshare$rideable_type) 
filtered_capital_bikeshare$member_casual <- as.factor(filtered_capital_bikeshare$member_casual)
filtered_capital_bikeshare$start_time_period <- as.factor(filtered_capital_bikeshare$start_time_period)
filtered_capital_bikeshare$day_type <- as.factor(filtered_capital_bikeshare$day_type)
```

*Removing outliers*
```{r}
#for distance_miles
z_scores_DM <- scale(filtered_capital_bikeshare$distance_miles)
outliers <- which(abs(z_scores_DM) > 3)
capital_bikeshare_clean <- filtered_capital_bikeshare[-outliers, ]

# for ride_duration_mins
z_scores_RDM <- scale(capital_bikeshare_clean$ride_duration_mins)
outliers <- which(abs(z_scores_RDM) > 2)
capital_bikeshare_clean <- capital_bikeshare_clean[-outliers, ]

```

After cleaning the dataset the number of rows are `r nrow(capital_bikeshare_clean)`.

# Visualizations:
*1. Frequency of electric bike versus classic bike rentals*
```{r}

bike_type_counts <- table(capital_bikeshare_clean$rideable_type)

bike_type_df <- as.data.frame(bike_type_counts)

# Create a bar plot to visualize the frequency of bike rentals by type
ggplot(bike_type_df, aes(x = Var1, y = Freq, fill = Var1)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Frequency of Electric Bike vs Classic Bike Rentals",
       x = "Bike Type",
       y = "Frequency of Rentals") + 
  scale_fill_manual(values = c("electric_bike" = "violet", "classic_bike" = "lightblue")) +  
  scale_y_continuous(labels = scales::comma) +
  theme(legend.position = "none")
```

- The higher frequency of electric bike rentals compared to classic bikes in the dataset indicates a preference for electric bikes. 


*2. Distribution of duration rides on electric bike versus classic bike*

```{r}
ggplot(capital_bikeshare_clean, aes(x = ride_duration_mins, fill = rideable_type)) +
  geom_density(alpha = 0.9) +  # Alpha sets the transparency for overlapping areas
  labs(title = "Distribution of Ride Durations for Electric vs Classic Bikes",
       x = "Ride Duration (minutes)",
       y = "Density") +
  scale_fill_manual(values = c("electric_bike" = "red", "classic_bike" = "green")) +  
  theme_minimal()

ggplot(capital_bikeshare_clean, aes(x = ride_duration_mins, fill = rideable_type)) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.7, position = "dodge") +
  labs(
    title = "Distribution of Ride Duration by Bike Type",
    x = "Ride Duration (Minutes)",
    y = "Frequency"
  ) +
  scale_fill_manual(values = c("electric_bike" = "pink", "classic_bike" = "red")) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

```
Box plot
```{r}
ggplot(capital_bikeshare_clean, aes(x = rideable_type, y = ride_duration_mins, fill = rideable_type)) +
  geom_boxplot() +
  labs(title = "Boxplot of Ride Durations for Electric vs Classic Bikes",
       x = "Bike Type",
       y = "Ride Duration (minutes)") +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "green")) +
  theme_minimal()
```



*3. Frequency of user types*

```{r}
user_type_counts <- table(capital_bikeshare_clean$member_casual)

user_type_df <- as.data.frame(user_type_counts)

ggplot(user_type_df, aes(x = Var1, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of User Types (Member vs Casual)",
       x = "User Type",
       y = "Frequency") +
  scale_fill_manual(values = c("member" = "blue", "casual" = "green")) +  
  theme_minimal()
```



Pie chart:
```{r}
ggplot(user_type_df, aes(x = "", y = Freq, fill = User_Type)) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar(theta = "y") +  # Convert the bar chart into a pie chart
  labs(title = "Proportion of User Types (Member vs Casual)") +
  scale_fill_manual(values = c("member" = "blue", "casual" = "green")) +
  theme_void() 
```

*4. User Types by Rideable Type*
```{r}
user_bike_counts <- capital_bikeshare_clean %>%
  group_by(member_casual, rideable_type) %>%
  summarise(Freq = n())  

ggplot(user_bike_counts, aes(x = rideable_type, y = Freq, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "User Type Distribution by Rideable Type",
       x = "Rideable Type (Bike Type)",
       y = "Frequency of Rides") +
  scale_fill_manual(values = c("member" = "pink", "casual" = "yellow")) +  
  theme_minimal()
```

```{r}
ggplot(user_bike_counts, aes(x = rideable_type, y = Freq, fill = member_casual)) +
  geom_bar(stat = "identity", position = "fill") +  
  labs(title = "Proportional User Type Distribution by Rideable Type",
       x = "Rideable Type (Bike Type)",
       y = "Proportion of Rides") +
  scale_fill_manual(values = c("member" = "pink", "casual" = "violet")) +  
  theme_minimal()
```

*5. Member vs Casual bike usage on weekdays vs weekends*
```{r}
ggplot(capital_bikeshare_clean, aes(x = day_type, fill = interaction(member_casual, rideable_type))) +
  geom_bar(position = "dodge") + # Dodged bars for separate member_casual and rideable_type combinations
  labs(title = "Rides by Day Type (Weekday vs Weekend)",
       x = "Day Type",
       y = "Count",
       fill = "Member Type & Rideable Type") +
  theme_minimal()
```

# Predict the rider's class based on trip characteristics
```{r}
riderLogit01 <- glm(member_casual ~ ride_duration_mins, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit01)

riderLogit02 <- glm(member_casual ~ rideable_type + ride_duration_mins, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit02)

riderLogit03 <- glm(member_casual ~ rideable_type + ride_duration_mins + day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit03)

riderLogit04 <- glm(member_casual ~ rideable_type + distance_miles + ride_duration_mins + day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit04)

riderLogit05 <- glm(member_casual ~ rideable_type * ride_duration_mins, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit05)

riderLogit06 <- glm(member_casual ~ rideable_type * ride_duration_mins * day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit06)

riderLogit07 <- glm(member_casual ~ rideable_type * distance_miles * ride_duration_mins * day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit07)

riderLogit08 <- glm(member_casual ~ rideable_type * distance_miles * ride_duration_mins * day_type * start_time_period, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit08)

riderLogit09 <- glm(member_casual ~ rideable_type + distance_miles + ride_duration_mins + start_time_period + day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit09)

riderLogit10 <- glm(member_casual ~ rideable_type * ride_duration_mins + distance_miles + start_time_period + day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit10)

riderLogit11 <- glm(member_casual ~ distance_miles + rideable_type * ride_duration_mins + day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit11)
```
AIC10 = 847467
AIC11 = 849314



ANOVA to compare models and select the best
```{r}
anova_result <- anova(riderLogit01, riderLogit02, riderLogit03, riderLogit04, riderLogit05, riderLogit06, riderLogit07, riderLogit08, riderLogit09, test = "Chisq")
print(anova_result)

anova_result <- anova(riderLogit04, riderLogit09, test = "Chisq")
print(anova_result)

anova_result <- anova(riderLogit09, riderLogit10, test = "Chisq")
print(anova_result)

anova_result <- anova(riderLogit04, riderLogit11, test = "Chisq")
print(anova_result)

anova_result <- anova(riderLogit11, riderLogit10, test = "Chisq")
print(anova_result)

anova_result <- anova(riderLogit04, riderLogit10, test = "Chisq")
print(anova_result)
```

```{r}

vif(riderLogit06)
vif(riderLogit04)
vif(riderLogit09)
vif(riderLogit10)
vif(riderLogit11)
```

Model selection started with duration (ride_duration_mins) only and then added bike type (rideable_type) because the decision tree indicated these were the most important variables (see below). Because this data doesn't have many variables, and nearly all were statistically significant in the full model, we elected a stepwise approach to model building, and used ANOVA to compare model quality. Initially Models 4 and 6 had the best reduction in residual deviance for minimal change in degrees of freedom that was statistically significant. However the VIF for model 6 indicated a significant amount of collinearity. So then we compared model 4 to the full model (9), which showed a smaller but still statistically significant reduction to residual deviance. However this model increased the complexity by 7 degrees of freedom. We tried 2 other versions of the model, incorporating the interaction terms between duration and bike type, but with and without the time-of-day (start_time_period). Model10(riderLogit10) including all variables and the interaction between duration and bike type is a better fit when compared via ANOVA to the model without time-of-day, with a statistically significant reduction in residual deviance.

Model 10 - member_casual ~ rideable_type * ride_duration_mins + distance_miles + start_time_period + day_type
```{r}
riderLogit10 <- glm(member_casual ~ rideable_type * ride_duration_mins + distance_miles + start_time_period + day_type, data = capital_bikeshare_clean, family = "binomial")
summary(riderLogit10)
```

The intercept is log-odds 1.52, the log-odds of being a member when the bike is a classic, ride duration and distance are 0, it is the afternoon, and it is a weekday.
Electric bike (verus classic bike) reduces the log-odds of being a member by log-odds 0.631.
Each additional minute of ride duration decrease the log-odds of being a member by 0.062.
Each additional mile of distance ridden increases the log-odds of being a member b 0.12
Evening, Late evening, Mid-morning, Morning increase the log-odds of being a member compared to afternoon. 
Early morning, midday, and Night are not statistically significant at alpha = 0.05.
Weekend rides reduce the likelihood of being a member by log-odds 0.42.
The interaction term indicates that for each additional minute of ride duration on an electric bike, the negative effect of ride duration (see above) on the log-odds of being a member is reduced by 0.04. This means longer rides on electric bikes are less strongly associated with with casual riders on classic bikes.



```{r}
exp(coef(riderLogit10))
```
When the log-odds are exponentiated, we can use the odds ratios (OR) to better understand the impact of the variables.
The intercept is 4.559 meaning the odds of being a member are 4.56 times higher than being a casual user when the bike is a classic, it is a weekday afternoon, and ride duration and distance are 0.
The odds ratio for electric bike is 0.541, meaning riders on electric bikes are 46% (1-0.541) less likely to be members compared to classic bikes.
The odds ratio for duraiton is 0.94, meaning each additional minute of ride duration decreases the odds of being a member by 6% (1-0.94).
The odds ratio for distance is 1.127, meaning each additional mile of distance increases the odds of being a member by 13%.
The odds ratio for morning rides.


```{r}
predicted_probs <- predict(riderLogit10, type = "response")

# Classify based on a threshold of 0.5
predicted_classes <- ifelse(predicted_probs > 0.5, "member", "casual")
predicted_classes <- factor(predicted_classes, levels = c("member", "casual"))
actual_classes <- factor(capital_bikeshare_clean$member_casual, levels = c("member", "casual"))

# Create the confusion matrix
conf_matrix <- confusionMatrix(predicted_classes, actual_classes)

# Print the confusion matrix and statistics
print(conf_matrix)

# Extract accuracy
accuracy <- conf_matrix$overall["Accuracy"]
print(paste("Accuracy:", round(accuracy, 4)))
```

Logit 10 Accuracy = 0.693


```{r}
predicted_probs10 <- predict(riderLogit10, newdata = capital_bikeshare_clean, type = "response")
roc_curve10 <- roc(response = capital_bikeshare_clean$member_casual, predicted_probs10, levels = c("casual", "member"))
plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)

auc_value10 <- auc(roc_curve10)
print(paste("AUC (ModelMetrics):", round(auc_value10, 4)))

```

AUC10 = 0.6337



# Model to predict the duration of the ride based on trip characteristics and type of rider.

```{r}

DurationModel <- rpart(ride_duration_mins ~ rideable_type + member_casual + distance_miles + start_time_period, data = capital_bikeshare_clean, method = "anova")

```
*Summary of Duration Model*
```{r}
print(summary(DurationModel))
```



```{r}
rpart.plot(DurationModel, type = 3, extra = 1, main = "Regression Tree for Ride Duration Prediction")

```
Regression tree to predict ride duration in minutes based on rideable_type, member_casual, distance_miles, and start_time_period.


- The root node contains all observations, with an average ride duration of 12.7 minutes.

- The first major split occurs at 1.55 miles for distance_miles. This results in two main branches:

  - For distances less than 1.55 miles, the average ride duration is 9.86 minutes.
  - For distances greater than or equal to 1.55 miles, the average ride duration increases to 20.7 minutes.
  
Key Predictors:

Distance is the primary factor influencing ride duration, with significant splits at 0.871 miles and 2.46 miles.
Rideable type and member status also significantly affect ride duration:
Members tend to have shorter ride durations than casual riders, especially for shorter distances.
Electric bikes are associated with shorter ride durations compared to classic bikes.


For trips with distance_miles < 0.871 and member status = member, the ride duration averages 6.96 minutes.
For casual riders with distance_miles < 0.871, the average ride duration is 10.90 minutes.
For trips with distance_miles ≥ 0.871 and < 2.46, the average duration is 12.6 minutes. Here, electric bike rides average 11.0 minutes, and classic bike rides average 15.2 minutes.
For longer trips with distance_miles ≥ 2.46, the ride duration increases further to 25.8 minutes, with electric bike rides averaging 23.6 minutes and classic bike rides averaging 32.4 minutes.
Model Performance:

The model effectively splits based on the most significant predictors, with the tree having 8 terminal nodes.

```{r}

# Build a tree for member_casual
Member_casual_tree <- rpart(member_casual~ rideable_type + distance_miles + ride_duration_mins + start_time_period + day_type, data = capital_bikeshare_clean, method = "class")

# Plot the tree
tree_vis <- rpart.plot(Member_casual_tree, type = 3, extra = 1, main = "Classification Tree for Member vs Casual Prediction")

# Print the tree
tree_vis

# Print the summary of the tree
printcp(Member_casual_tree)

# K-fold 
fold <- floor(runif(nrow(capital_bikeshare_clean),1,11)) 
  capital_bikeshare_clean$fold <- fold
## the test set is just the first fold 
test.set <- capital_bikeshare_clean[capital_bikeshare_clean$fold == 1,] 
##exclude the first fold from the data here 
train.set <- capital_bikeshare_clean[capital_bikeshare_clean$fold != 1,]
treefit_fold <- rpart(member_casual~ rideable_type + distance_miles + ride_duration_mins + start_time_period, day_type, data = capital_bikeshare_clean, method = "class")

fancyRpartPlot(treefit_fold)

pred <- predict(treefit_fold, test.set) 
plot(pred,test.set$member_casual, main="Predicted vs. Observed Salary", xlab="Tree Predicted Salary", ylab="Observed Salary")
treefit_fold_error = sum(pred-test.set$member_casual)^2
treefit_fold_error <- sum((pred - test.set$member_casual)^2)
pred <- pred[, 2]  # Convert matrix to vector
treefit_fold_error <- sum((pred - test.set$member_casual)^2)
print(treefit_fold_error)





```


```{r}
library(rattle)

fancyRpartPlot(DurationModel)
```


```{r}
print(DurationModel$cptable)
```
From this table:

The lowest xerror occurs at cp = 0.0100, where xerror is 0.689.
However, the xerror is very close between cp = 0.0102 (0.697) and cp = 0.0100 (0.689). The difference is very small, suggesting that the improvement in error with cp = 0.0100 is marginal.
```{r}
plotcp(DurationModel)
```

Cross- validation:

```{r, results="markup"}
fold <- floor(runif(nrow(capital_bikeshare_clean),1,11)) 
capital_bikeshare_clean$fold <- fold
## the test set is just the first fold 
test.set <- capital_bikeshare_clean[capital_bikeshare_clean$fold == 1,] 
##exclude the first fold from the data here 
train.set <- capital_bikeshare_clean[capital_bikeshare_clean$fold != 1,]
DurationModel_fold <- rpart(ride_duration_mins ~ rideable_type + member_casual +distance_miles + start_time_period, data=train.set, method="anova" )
fancyRpartPlot(DurationModel_fold)
```

```{r}
pred <- predict(DurationModel_fold, test.set) 
plot(pred,test.set$ride_duration_mins, main="Predicted vs. Observed ride_duration_mins", xlab="Tree Predicted ride_duration_mins", ylab="Observed ride_duration_mins")
treefit_fold_error = sum(pred-test.set$ride_duration_mins)^2
print(treefit_fold_error)
```
The tree fold error is too large.

now considering only `rideable_type`, `member_casual`, `start_time_period`.
```{r, results="markup"}
DurationModel_fold2 <- rpart(ride_duration_mins ~ rideable_type + member_casual  + start_time_period, data=train.set, method="anova" )
fancyRpartPlot(DurationModel_fold2)
```

```{r}
pred2 <- predict(DurationModel_fold2, test.set) 
plot(pred2,test.set$ride_duration_mins, main="Predicted vs. Observed ride_duration_mins", xlab="Tree Predicted ride_duration_mins", ylab="Observed ride_duration_mins")
treefit_fold_error2 = sum(pred2-test.set$ride_duration_mins)^2
print(treefit_fold_error2)
```


The tree fold error is reduced  but still is large. But given that the test data is `r nrow(test.set)` the average error per prediction is `r treefit_fold_error2/nrow(test.set)`.

```{r}
summary(DurationModel_fold2)
```

To predict for a new data.

```{r}
new_data <- data.frame(
  rideable_type = factor("electric_bike"),
  member_casual = factor("casual"),
  start_time_period = factor("Morning")
)

# Predict the ride duration
predicted_duration <- predict(DurationModel_fold2, new_data)
predicted_duration
```



```{r}
rideable_typeLogit <- glm(rideable_type ~ member_casual + distance_miles + ride_duration_mins + start_time_period, data = capital_bikeshare_clean, family = "binomial")

summary(rideable_typeLogit)
```
The logistic regression model shows that several factors significantly influence the `rideable_type.` The coefficient for `member_casualmember` indicates that `members` are less likely to use the specified `rideable_type` compared to `casual` users. `Distance_miles` has a positive impact, suggesting that longer rides are more likely to use this `rideable_type`, while `ride_duration_mins` negatively affects the likelihood. Among the time periods, Early Morning, Night, and Late Evening show higher log-odds compared to other periods like Morning, Mid-Morning, and Evening, which have lower log-odds. The model’s residual deviance (899759) is much lower than the null deviance (933757), indicating a good fit. 

```{r}
rideable_typeModel <- rpart(rideable_type ~ ride_duration_mins+member_casual + distance_miles + start_time_period, data = capital_bikeshare_clean, method = "class")
print(summary(rideable_typeModel))
```

```{r}
rpart.plot(rideable_typeModel, type = 3, extra = 1, main = "Classification Tree for Rideable type Prediction")
```
```{r}
print(rideable_typeModel$cptable)
```

```{r}
plotcp(rideable_typeModel)
```

```{r, results="markup"}
fold <- floor(runif(nrow(capital_bikeshare_clean),1,11)) 
capital_bikeshare_clean$fold <- fold
## the test set is just the first fold 
test.set <- capital_bikeshare_clean[capital_bikeshare_clean$fold == 1,] 
##exclude the first fold from the data here 
train.set <- capital_bikeshare_clean[capital_bikeshare_clean$fold != 1,]
rideable_typeModel_fold <- rpart(rideable_type ~ ride_duration_mins+member_casual + distance_miles + start_time_period, data=train.set, method="class" )
fancyRpartPlot(rideable_typeModel_fold)
```


```{r}
pred <- predict(rideable_typeModel_fold, test.set) 
plot(pred,test.set$rideable_type, main="Predicted vs. Observed rideable_type", xlab="Tree classified rideable_type", ylab="Observed rideable_type")
treefit_fold_error = sum(pred-test.set$rideable_type)^2
print(treefit_fold_error)
```


```{r}
library(caret) 
cm = confusionMatrix( predict(rideable_typeModel, type = "class"), reference = capital_bikeshare_clean[, "rideable_type"] )
print('Overall: ')
cm$overall
print('Class: ')
cm$byClass

```

