---
title: "Analysis and modelling on the Capital Bikeshare data"
author: "[Anusha Umashankar](https://github.com/Anusha-raju)| [Aidan Carlisle](https://github.com/acarlisle8)| [Rachel Thomas](https://github.com/releered)| [Sayan Patra](https://github.com/Sayanpatraa) - Group 8 "
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<style>

h1 { color: #800000; }  

h2 { color: #000080; } 
h3 { color: #004d4d; } 

</style>

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(geosphere)
library(tidygeocoder)
```



```{r}
file_path <- file.path(getwd(), "202410-capitalbikeshare-tripdata.csv")
capital_bikeshare <- read.csv(file_path, header = TRUE)
```

The dataset contains the following attributes.
```{r}
colnames(capital_bikeshare)
```

# Preprocessing:

1. 
There are data points where information like start_station_name & id , end_station_name & id , and end_lat and end_lng are missing. 
`r colSums(is.na(capital_bikeshare))`.

Dropping the `r colSums(is.na(capital_bikeshare))["end_lat"]` rows as they consitute only `r colSums(is.na(capital_bikeshare))["end_lat"]/nrow(capital_bikeshare)*100` percent of the dataset.

```{r}
capital_bikeshare <- capital_bikeshare %>%
  filter(!is.na(end_lat))
```


2. Reverse Geocoding: Converting Latitude and Longitude Coordinates to Addresses
```{r}
#capital_bikeshare$start_addy<-reverse_geocode(capital_bikeshare, lat = "start_lat", long = "start_lng")
#capital_bikeshare$end_addy<-reverse_geocode(capital_bikeshare, lat = "end_lat", long = "end_lng")
```

3. `Distance (in miles)` is derived using the latitude and longitude coordinates of the start and end points. The `distVincentySphere()` function from the `geosphere` package in R is used to compute the geodesic distance, factoring in the Earth's curvature.

```{r}
# Calculate the distance for each row and add it as a new column (in miles)
capital_bikeshare$distance_miles <- mapply(function(lat1, lon1, lat2, lon2) {
  distVincentySphere(c(lon1, lat1), c(lon2, lat2)) / 1000 * 0.621371  
}, capital_bikeshare$start_lat, capital_bikeshare$start_lng, capital_bikeshare$end_lat, capital_bikeshare$end_lng)
```


4. Calculated the total ride time by subtracting `started_at` from `ended_at.`
```{r}
capital_bikeshare$started_at <- as.POSIXct(capital_bikeshare$started_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ended_at <- as.POSIXct(capital_bikeshare$ended_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ride_duration_mins <- as.numeric(difftime(capital_bikeshare$ended_at, capital_bikeshare$started_at, units = "mins"))
```


5. Categorizing Rides by Time Period.
```{r}
get_time_period <- function(time) {
  hour <- as.numeric(format(time, "%H"))
  
  if (hour >= 4 & hour < 6) {
    return("Early Morning")
  } else if (hour >= 6 & hour < 9) {
    return("Morning")
  } else if (hour >= 9 & hour < 12) {
    return("Mid-Morning")
  } else if (hour >= 12 & hour < 13) {
    return("Midday")
  } else if (hour >= 13 & hour < 17) {
    return("Afternoon")
  } else if (hour >= 17 & hour < 19) {
    return("Evening")
  } else if (hour >= 19 & hour < 21) {
    return("Late Evening")
  } else {
    return("Night")
  }
}


capital_bikeshare <- capital_bikeshare %>%
  mutate(
    start_time_period = sapply(started_at, get_time_period),
  )
```


**The following attributes are considered for further Exploratory Data Analysis (EDA):**

```{r}
#filtered_capital_bikeshare <- capital_bikeshare %>% select(rideable_type,member_casual,distance_miles,ride_duration_mins, start_time_period, start_addy, end_addy)
filtered_capital_bikeshare <- capital_bikeshare %>% select(rideable_type,member_casual,distance_miles,ride_duration_mins, start_time_period)
```

*Factorizing `rideable_type`, `member_casual` and `start_time_period` variables:*
```{r}
filtered_capital_bikeshare$rideable_type <- as.factor(filtered_capital_bikeshare$rideable_type) 
filtered_capital_bikeshare$member_casual <- as.factor(filtered_capital_bikeshare$member_casual)
filtered_capital_bikeshare$start_time_period <- as.factor(filtered_capital_bikeshare$start_time_period)
```

*Removing outliers*
```{r}
#for distance_miles
z_scores_DM <- scale(filtered_capital_bikeshare$distance_miles)
outliers <- which(abs(z_scores_DM) > 3)
capital_bikeshare_clean <- filtered_capital_bikeshare[-outliers, ]

# for ride_duration_mins
z_scores_RDM <- scale(capital_bikeshare_clean$ride_duration_mins)
outliers <- which(abs(z_scores_RDM) > 2)
capital_bikeshare_clean <- capital_bikeshare_clean[-outliers, ]

```

After cleaning the dataset the number of rows are `r nrow(capital_bikeshare_clean)`.

# Visualizations:
*1. Frequency of electric bike versus classic bike rentals*
```{r}

bike_type_counts <- table(capital_bikeshare_clean$rideable_type)

bike_type_df <- as.data.frame(bike_type_counts)

# Create a bar plot to visualize the frequency of bike rentals by type
ggplot(bike_type_df, aes(x = Var1, y = Freq, fill = Var1)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Frequency of Electric Bike vs Classic Bike Rentals",
       x = "Bike Type",
       y = "Frequency of Rentals") + 
  scale_fill_manual(values = c("electric_bike" = "violet", "classic_bike" = "lightblue")) +  
  theme_minimal()
```

- The higher frequency of electric bike rentals compared to classic bikes in the dataset indicates a preference for electric bikes. 


*2. Distribution of duration rides on electric bike versus classic bike*

```{r}
ggplot(capital_bikeshare_clean, aes(x = ride_duration_mins, fill = rideable_type)) +
  geom_density(alpha = 0.9) +  # Alpha sets the transparency for overlapping areas
  labs(title = "Distribution of Ride Durations for Electric vs Classic Bikes",
       x = "Ride Duration (minutes)",
       y = "Density") +
  scale_fill_manual(values = c("electric_bike" = "red", "classic_bike" = "green")) +  
  theme_minimal()
```
Box plot
```{r}
ggplot(capital_bikeshare_clean, aes(x = rideable_type, y = ride_duration_mins, fill = rideable_type)) +
  geom_boxplot() +
  labs(title = "Boxplot of Ride Durations for Electric vs Classic Bikes",
       x = "Bike Type",
       y = "Ride Duration (minutes)") +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "green")) +
  theme_minimal()
```



*3. Frequency of user types*

```{r}
user_type_counts <- table(capital_bikeshare_clean$member_casual)

user_type_df <- as.data.frame(user_type_counts)

ggplot(user_type_df, aes(x = Var1, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of User Types (Member vs Casual)",
       x = "User Type",
       y = "Frequency") +
  scale_fill_manual(values = c("member" = "blue", "casual" = "green")) +  
  theme_minimal()
```



Pie chart:
```{r}
ggplot(user_type_df, aes(x = "", y = Freq, fill = User_Type)) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar(theta = "y") +  # Convert the bar chart into a pie chart
  labs(title = "Proportion of User Types (Member vs Casual)") +
  scale_fill_manual(values = c("member" = "blue", "casual" = "green")) +
  theme_void() 
```

*4. User Types by Rideable Type*
```{r}
user_bike_counts <- capital_bikeshare_clean %>%
  group_by(member_casual, rideable_type) %>%
  summarise(Freq = n())  

ggplot(user_bike_counts, aes(x = rideable_type, y = Freq, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "User Type Distribution by Rideable Type",
       x = "Rideable Type (Bike Type)",
       y = "Frequency of Rides") +
  scale_fill_manual(values = c("member" = "pink", "casual" = "yellow")) +  
  theme_minimal()
```

```{r}
ggplot(user_bike_counts, aes(x = rideable_type, y = Freq, fill = member_casual)) +
  geom_bar(stat = "identity", position = "fill") +  
  labs(title = "Proportional User Type Distribution by Rideable Type",
       x = "Rideable Type (Bike Type)",
       y = "Proportion of Rides") +
  scale_fill_manual(values = c("member" = "pink", "casual" = "violet")) +  
  theme_minimal()
```

# Predict the rider's class based on trip characteristics
```{r}
riderLogit <- glm(member_casual ~ rideable_type + distance_miles + ride_duration_mins + start_time_period, data = capital_bikeshare_clean, family = "binomial")

summary(riderLogit)
```

Electric bike reduces the log-odds of being a member compared to a classic bike.
Electric scooter strongly reduces the log-odds of being a member compared to the classic bike.
Positive distance coefficient indicates longer distances increase the likelihood of being a member.
Negative duration coefficition indicates that longer rides decrease the likelihood of being a member.
Evening, Late evening, Mid-morning, and Morning increase the log-odds of being a member. Night decreases the log odds of being a member. 
Early morning and midday are not statistically significant.

```{r}
exp(coef(riderLogit))
```

The odds of being a member (the intercept) is 2.78 times higher than being a casual rider.
The OR for electric bike is 0.88, meaning the odds of being a member are about 12% lower (1-0.88) for electric bike users compared to classic bike.
The OR for electric scooter is 0.07, meaning the odds of bing a member are about 93% lower (1-0.07) for electric scooter users compared to classic bike.
The OR for distance is 1.2126, meaning each additional mile traveled increased the odds of being a member by 21.26%.
The OR for duration is 0.9570, meaning each additional minute of ride duration decreases the odds of the being a member by 4.3% (1-0.9570).
The OR for early morning is 0.9767, meaning rides starting during that time period have a 2.33% (1-0.9767) lower odds of being a member compared to afternoon (the reference period).
The OR for evening is 1.1349, meaning rides starting in the evening have 13.49% higher odds of being a member compared to the afternoon.
The OR for Late evening is 1.1025, meaning rides starting in the late evening have a 10.25% higher odds of being a member compared to the afternoon.
The OR for mid-morning is 1.0743, meaning these rides have 7.43% higher odds of being a member.
The OR for Midday is 1.0063, meaning these rides have 0.63% higher odds of being a member.
The OR for morning is 1.6012, meaning morning rides have 60% higher odds of being a member.
The OR for Night is 0.9571, meaning night rides have 4.29% lower odds of being a member.

Members are more likely to use classic bikes. take longer rides, ride for shorter duration, and are most associated with morning and evening rides.

```{r}
predicted <- ifelse(predict(riderLogit, type = "response") > 0.5, "member", "casual")
conf_matrix <- table(Predicted = predicted, Actual = capital_bikeshare_clean$member_casual)
TP <- conf_matrix[2, 2] 
TN <- conf_matrix[1, 1]  
FP <- conf_matrix[1, 2]  
FN <- conf_matrix[2, 1]  
accuracy <- (TP + TN) / (TP + TN + FP + FN)  
print(paste("Accuracy:", round(accuracy, 4)))
```

This model's accuracy is `r round(accuracy, 4)))` which is not great.

```{r}
library(car)
vif(riderLogit)
```

The VIF has all variables at nearly 1 which means there is no appreciable multicollinearity.

```{r}
riderLogit_stepwise <- step(
  riderLogit, 
  direction = "backward"
)
```

The AIC is increased by removing any of the predictors, much worse if you remove duration, and much, much worse if you remove rideable_type. So these predictors all contribute meaningfully to the model.

riderLogit_refined <- glm(
  member_casual ~ rideable_type + distance_miles + ride_duration_mins + 
    start_time_periodEvening + start_time_periodLate Evening +
    start_time_periodMid-Morning + start_time_periodMorning + 
    start_time_periodNight,
  family = "binomial",
  data = capital_bikeshare_clean
)
