---
title: "Analysis and modelling on the Capital Bikeshare data"
author: "[Anusha Umashankar](https://github.com/Anusha-raju)| [Aidan Carlisle](https://github.com/acarlisle8)| [Rachel Thomas](https://github.com/releered)| [Sayan Patra](https://github.com/Sayanpatraa) - Group 8 "
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<style>

h1 { color: #800000; }  

h2 { color: #000080; } 
h3 { color: #004d4d; } 

</style>

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(geosphere)
```


```{r}
capital_bikeshare <- read.csv("~/workspace/Data science/Capital-Bikeshare/202410-capitalbikeshare-tripdata.csv", header = TRUE)
```

The dataset contains the following attributes.
```{r}
colnames(capital_bikeshare)
```

# Preprocessing:

1. `Distance (in miles)` is derived using the latitude and longitude coordinates of the start and end points. The `distVincentySphere()` function from the `geosphere` package in R is used to compute the geodesic distance, factoring in the Earth's curvature.

```{r}
# Calculate the distance for each row and add it as a new column (in miles)
capital_bikeshare$distance_miles <- mapply(function(lat1, lon1, lat2, lon2) {
  distVincentySphere(c(lon1, lat1), c(lon2, lat2)) / 1000 * 0.621371  
}, capital_bikeshare$start_lat, capital_bikeshare$start_lng, capital_bikeshare$end_lat, capital_bikeshare$end_lng)
```


2. Calculated the total ride time by subtracting `started_at` from `ended_at.`
```{r}
capital_bikeshare$started_at <- as.POSIXct(capital_bikeshare$started_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ended_at <- as.POSIXct(capital_bikeshare$ended_at, format="%Y-%m-%d %H:%M:%OS")
capital_bikeshare$ride_duration_mins <- as.numeric(difftime(capital_bikeshare$ended_at, capital_bikeshare$started_at, units = "mins"))
```


3. Categorizing Rides by Time Period.
```{r}
get_time_period <- function(time) {
  hour <- as.numeric(format(time, "%H"))
  
  if (hour >= 4 & hour < 6) {
    return("Early Morning")
  } else if (hour >= 6 & hour < 9) {
    return("Morning")
  } else if (hour >= 9 & hour < 12) {
    return("Mid-Morning")
  } else if (hour >= 12 & hour < 13) {
    return("Midday")
  } else if (hour >= 13 & hour < 17) {
    return("Afternoon")
  } else if (hour >= 17 & hour < 19) {
    return("Evening")
  } else if (hour >= 19 & hour < 21) {
    return("Late Evening")
  } else {
    return("Night")
  }
}


capital_bikeshare <- capital_bikeshare %>%
  mutate(
    start_time_period = sapply(started_at, get_time_period),
  )
```

**The following attributes are considered for further Exploratory Data Analysis (EDA):**

```{r}
filtered_capital_bikeshare <- capital_bikeshare %>% select(rideable_type,member_casual,distance_miles,ride_duration_mins, start_time_period)
```





#Preprocessing

oct2024_capbike <- X202410_capitalbikeshare_tripdata

summary(oct2024_capbike)

str(oct2024_capbike)
unique(oct2024_capbike$rideable_type)
unique(oct2024_capbike$member_casual)

# Factorize categorical variables
oct2024_capbike$rideable_type <- factor(oct2024_capbike$rideable_type)
oct2024_capbike$member_casual <- factor(oct2024_capbike$member_casual)

str(oct2024_capbike)

#Compute duration of each ride
oct2024_capbike$duration_seconds <- as.numeric(difftime(oct2024_capbike$ended_at, oct2024_capbike$started_at, units = "secs"))

oct2024_capbike$duration_minutes <- oct2024_capbike$duration_seconds / 60

str(oct2024_capbike)
summary(oct2024_capbike)

#Compute distance traveled for each ride

install.packages("sf")
library(sf)

sum(is.na(oct2024_capbike$end_lat))
sum(is.na(oct2024_capbike$end_lng))

capbike_clean <- oct2024_capbike[!is.na(oct2024_capbike$end_lat) & !is.na(oct2024_capbike$end_lng), ]

sum(is.na(capbike_clean$end_lat))
sum(is.na(capbike_clean$end_lng))
sum(is.na(oct2024_capbike$start_lat))
sum(is.na(oct2024_capbike$start_lng))

capbike_sf <- st_as_sf(capbike_clean, coords = c("start_lng", "start_lat"), crs = 4326)
end_points <- st_as_sf(capbike_clean, coords = c("end_lng", "end_lat"), crs = 4326)

capbike_clean$distance_meters <- st_distance(capbike_sf, end_points, by_element = TRUE)

capbike_clean$distance_km <- as.numeric(capbike_clean$distance_meters) / 1000

str(capbike_clean)
summary(capbike_clean)

#Assess outliers
capbike_clean <- outlierKD2(capbike_clean, duration_minutes, rm = TRUE, boxplt = TRUE, qqplt = TRUE)

capbike_clean <- outlierKD2(capbike_clean, distance_km, rm = TRUE, boxplt = TRUE, qqplt = TRUE)

# The outliers for duration are crazy extreme (max 1500 minutes) - vs without the outliers max is 34. Removing the outliers removes ~54.5k obs, which is a lot, but still leaves 670279 obs which I think is more than enough. So my vote is to remove the outliers. I applied the outlierKD2 here too - removed 31014 obs, but still have more than 600k which is more than enough. So I've made the executive decision to remove all of the NAs in duration_minutes and distance_km from the dataset (we can obviously undo if you guys disagree).

capbike_cleanest <- capbike_clean %>%
  filter(!is.na(distance_km) & !is.na(duration_minutes))


#EDA

# SMART1: Bar plot for frequency of bike type

ggplot(data = capbike_cleanest, aes(x = rideable_type)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Bike Types",
    x = "Type of Bike",
    y = "Frequency"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
  
  
# SMART2: Distribution of duration on ebike v classic

ggplot(capbike_cleanest, aes(x = duration_minutes, fill = rideable_type)) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.7, position = "dodge") +
  labs(
    title = "Distribution of Ride Duration by Bike Type",
    x = "Ride Duration (Minutes)",
    y = "Frequency"
  ) +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "orange")) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
  
ggplot(capbike_cleanest, aes(x = rideable_type, y = duration_minutes, fill = rideable_type)) +
  geom_boxplot() +
  labs(
    title = "Box Plot of Ride Duration by Bike Type",
    x = "Rideable Type",
    y = "Duration (Minutes)"
  ) +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "orange")) +  # Custom colors
  theme_minimal()



# SMART3: Distribution of distances ridde by bike type. 

ggplot(capbike_cleanest, aes(x = distance_km, fill = rideable_type)) +
  geom_histogram(binwidth = 0.3, color = "black", alpha = 0.7, position = "dodge") +
  labs(
    title = "Distribution of Ride Distance by Bike Type",
    x = "Ride Distance (km)",
    y = "Frequency"
  ) +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "orange")) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
  
ggplot(capbike_cleanest, aes(x = rideable_type, y = distance_km, fill = rideable_type)) +
  geom_boxplot() +
  labs(
    title = "Box Plot of Ride Distance by Bike Type",
    x = "Rideable Type",
    y = "Distance (km)"
  ) +
  scale_fill_manual(values = c("electric_bike" = "blue", "classic_bike" = "orange")) + 
  theme_minimal()
  
# SMART4: Frequency of user types

ggplot(data = capbike_cleanest, aes(x = member_casual)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of User Types",
    x = "Type of User",
    y = "Frequency"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()


  

